---
title: "Seasonality"
author: "Lajos Galambos"
format: html
---

## McDonald's (MCD) seasonality analysis

I hold MCD stock in my portfolio. Recently, it has made significant gains, and I am interested in understanding the seasonality of its returns. I will analyze the average monthly returns of MCD stock over the past 10 years to identify any seasonal patterns.

```{r}
#| echo: false
#| output: false
#| message: false
#| warning: false
#install.packages("tidyquant")
#install.packages("dplyr")
#install.packages("ggplot2")
library(tidyquant)
library(dplyr)
library(ggplot2)
```

I take data from Yahoo Finance for McDonald's (MCD) stock from March 10, 2000, to March 10, 2025. I calculate the monthly returns and group them by month to compute the average monthly return for each month.

```{r}
#| echo: false
#| message: false
#| warning: false
ticker <- "MCD"
start_date <- "2000-03-10"
end_date <- "2025-03-10"

mcd_data <- tq_get(ticker, from = start_date, to = end_date)

mcd_data <- mcd_data %>%
  mutate(Return = adjusted / lag(adjusted) - 1) %>%
  mutate(Year = format(date, "%Y"),
         Month = format(date, "%m")) 

monthly_seasonality <- mcd_data %>%
  group_by(Month) %>%
  summarise(Average_Return = mean(Return, na.rm = TRUE) * 100)

monthly_seasonality$Month <- factor(monthly_seasonality$Month, 
                                    levels = sprintf("%02d", 1:12), 
                                    labels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", 
                                               "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"))

ggplot(monthly_seasonality, aes(x = Month, y = Average_Return)) +
  geom_bar(stat = "identity", fill = "blue", alpha = 0.7) +
  theme_minimal() +
  labs(title = "McDonald's (MCD) Monthly Seasonality (Avg. Monthly Return)",
       x = "Month",
       y = "Average Monthly Return (%)") +
  theme(axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 12))
```
I also create a heatmap to visualize the seasonality of MCD stock over the past 10 years.

```{r}
#| echo: false
#| message: false
#| warning: false
seasonality_matrix <- mcd_data %>%
  group_by(Year, Month) %>%
  summarise(Average_Return = mean(Return, na.rm = TRUE) * 100) %>%
  ungroup()

seasonality_matrix$Month <- factor(seasonality_matrix$Month, 
                                   levels = sprintf("%02d", 1:12), 
                                   labels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", 
                                              "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"))

seasonality_matrix$Year <- factor(seasonality_matrix$Year, levels = rev(unique(seasonality_matrix$Year)))

ggplot(seasonality_matrix, aes(x = Month, y = Year, fill = Average_Return)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "red", mid = "white", high = "green", midpoint = 0) +
  labs(title = "McDonald's (MCD) Seasonality Matrix (Avg. Monthly Return %)",
       x = "Month",
       y = "Year",
       fill = "Avg Return (%)") +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 12, angle = 45, hjust = 1),
        axis.text.y = element_text(size = 9),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())
```


Finally, lets count the number of positive and negative monthly returns for each month to see if there is a pattern in the number of positive and negative months for each month.

```{r}
#| echo: false
#| message: false
#| warning: false

if (!requireNamespace("gt", quietly = TRUE)) install.packages("gt")

library(gt)
library(dplyr)
library(tidyr)

positive_negative_counts <- mcd_data %>%
  group_by(Year, Month) %>%
  summarise(Avg_Return = mean(Return, na.rm = TRUE) * 100) %>%
  ungroup() %>%
  mutate(Result = ifelse(Avg_Return > 0, "Positive", "Negative")) %>%
  group_by(Month, Result) %>%
  summarise(Count = n()) %>%
  tidyr::spread(Result, Count, fill = 0) %>%
  mutate(Total = Positive + Negative,
         Positive_Percentage = round((Positive / Total), 2))

positive_negative_counts$Month <- factor(positive_negative_counts$Month, 
                                         levels = sprintf("%02d", 1:12), 
                                         labels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", 
                                                    "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"))

positive_negative_counts %>%
  gt() %>%
  tab_header(
    title = "Historical Monthly Returns: Positive vs. Negative"
  ) %>%
  fmt_number(
    columns = c(Positive, Negative, Total),
    decimals = 0
  ) %>%
  fmt_percent(
    columns = Positive_Percentage,
    decimals = 0
  ) %>%
  data_color(
    columns = Positive_Percentage,
    colors = scales::col_numeric(
      palette = c("red", "yellow", "green"),
      domain = c(min(positive_negative_counts$Positive_Percentage), max(positive_negative_counts$Positive_Percentage))
    )
  ) %>%
  cols_label(
    Month = "Month",
    Positive = "Positive Months",
    Negative = "Negative Months",
    Total = "Total Years",
    Positive_Percentage = "Positive %"
  ) %>%
  tab_options(
    table.font.size = px(14),
    heading.title.font.size = px(18),
    column_labels.font.weight = "bold"
  )
```
## Remarks 

November has the highest percentage of positive months (96%), making it the most consistently strong month. June has the lowest percentage of positive months (32%), indicating it is the weakest. April (76%), March (69%), and July (68%) show strong positive trends, while January (48%), May (52%), and December (52%) are more balanced. This data highlights seasonal trends in historical market performance.


