---
title: "Turkey and Exchange Rate Overview"
author: "Lajos Galambos"
format: html
---

# Turkish Lira

The Turkish Lira has experienced serious devaluation in the last decade. Much of its devaluation (against major currencies) has been paralleled with the recent macroeconomic struggles of Turkey. Sky high inflation, global shocks, inappropriate economic policies, and political interference made the problems especially acute.

The Economist (a newspaper) in an article shed some light on Turkey's way out from the crisis ([Turkey’s long hard struggle with inflation](https://www.economist.com/europe/2024/10/10/turkeys-long-hard-struggle-with-inflation)). After mixed policy choices of the Turkish Central Bank, interest rates are high now, and the Economist claims that consumer prices are increasing in slower pace. The new monetary toolkit seems to be effective.

```{r}
#| echo: false
#| message: false
#| warning: false
#| output: false
library(quantmod)
library(ggplot2)
```

```{r}
#| echo: false
#| message: false
#| warning: false
#| output: false
# Get the Exchange Rate Data
symbol <- "TRY=X"

# 10 years historical data
start_date <- Sys.Date() - 365.25 * 10  # 10 years ago
end_date <- Sys.Date()  # today

# Get exchange rate data
getSymbols(symbol, src = "yahoo", from = start_date, to = end_date)
exchange_rate <- Cl(get(symbol))

exchange_rate_df <- data.frame(Date = index(exchange_rate), Close = as.numeric(exchange_rate))
```

## Interest Rates in Turkey and in the US

### Figure 1:

```{r}
#| echo: false
#| message: false
#| warning: false
# Plot exchange rate data
ggplot(exchange_rate_df, aes(x = Date, y = Close)) +
  geom_line(color = "#1c61b6") +
  ggtitle("USD/TRY Exchange Rate (Last 10 Years)") +
  xlab("Date") +
  ylab("Exchange Rate") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, color = "#666666", size = 14, face = "bold"),
        axis.title.x = element_text(color = "#666666", size = 12),
        axis.title.y = element_text(color = "#666666", size = 12),
        panel.border = element_rect(fill = NA, color = "#666666", size = 0.5, linetype = "solid"),
        panel.grid.major = element_line(size = 0.2, linetype = 'solid', color = "#cccccc"),
        panel.grid.minor = element_line(size = 0.2, linetype = 'solid', color = "#cccccc"))
```

```{r}
#| echo: false
#| message: false
#| warning: false
#| output: false
# Create a ratio
first_exchange_rate<- exchange_rate_df$Close[1]
last_exchange_rate <- tail(exchange_rate_df$Close, n = 1)
ratio <- last_exchange_rate / first_exchange_rate
print(first_exchange_rate)
print(last_exchange_rate)
print(ratio)
```

Figure 1 shows the historical exchange rate of 1 US Dollar in Turkish Lira units. It is clear that the Lira's devaluation was severe, for 1 Turkish Lira you could receive 15 times less USD dollars in 2024 than in 2014. Data is taken from [finance.yahoo.com](https://finance.yahoo.com/quote/TRY=X/).

|                                              |                         |
|----------------------------------------------|-------------------------|
| 2024 Exchange Rate Level                     | 2.2703 (TRY for 1 USD)  |
| 2014 Exchange Rate Level                     | 34.2743 (TRY for 1 USD) |
| Ratio of Recent and Old Exchange Rate Levels | 15.0964                 |

: Ratio of recent and old exchange rate

```{r}
#| echo: false
#| message: false
#| warning: false
#| output: false
# Interest Rate Data TURKEY
symbol <- "IRSTCI01TRM156N"
# IRSTCI01TRM156N represents the monthly national short-term interbank interest rate for Turkey. 

# Define start and end dates
start_date <- "2014-01-01"  # start of 2014
end_date <- Sys.Date()  # today

# Get interest rate data
getSymbols(symbol, src = "FRED", from = start_date, to = end_date)
interest_rate <- get(symbol)

# Convert to data frame for ggplot2
interest_rate_df <- data.frame(Date = index(interest_rate), InterestRate = as.numeric(interest_rate))
```

### Figure 2:

```{r}
#| echo: false
#| message: false
#| warning: false
# Plot interest rate data TURKEY
ggplot(interest_rate_df, aes(x = Date, y = InterestRate)) +
  geom_line(color = "#1c61b6") +
  ggtitle("Turkey Interest Rate (Since 2014)") +
  xlab("Date") +
  ylab("Interest Rate (%)") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, color = "#666666", size = 14, face = "bold"),
        axis.title.x = element_text(color = "#666666", size = 12),
        axis.title.y = element_text(color = "#666666", size = 12),
        panel.border = element_rect(fill = NA, color = "#666666", size = 0.5, linetype = "solid"),
        panel.grid.major = element_line(size = 0.2, linetype = 'solid', color = "#cccccc"),
        panel.grid.minor = element_line(size = 0.2, linetype = 'solid', color = "#cccccc"))
```

Figure 2 shows historical interest rates (short term, policy rate) in percentages for Turkey. The data contains numerous missing observations. Monetary policy in Turkey had reversal in strategy amid the COVID pandemic, recently got back again to the high interest rate environment. Data is taken from the [Federal Reserve Economic Data](https://fred.stlouisfed.org/series/IRSTCI01TRM156N).

### Figure 3:

```{r}
#| echo: false
#| message: false
#| warning: false
#| output: false
# Interest Rate Data US
symbol_US <- "IRSTCI01USM156N"

# Define start and end dates
start_date_US <- "2014-01-01"  # start of 2014
end_date_US <- Sys.Date()  # today

# Get interest rate data
getSymbols(symbol_US, src = "FRED", from = start_date_US, to = end_date_US)
interest_rate_US <- get(symbol_US)

# Convert to data frame for ggplot2
interest_rate_df_US <- data.frame(Date = index(interest_rate_US), InterestRate = as.numeric(interest_rate_US))
```

```{r}
#| echo: false
#| message: false
#| warning: false
# Plot interest rate data
ggplot(interest_rate_df_US, aes(x = Date, y = InterestRate)) +
  geom_line(color = "#1c61b6") +
  ggtitle("US Interest Rate (Since 2014)") +
  xlab("Date") +
  ylab("Interest Rate (%)") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, color = "#666666", size = 14, face = "bold"),
        axis.title.x = element_text(color = "#666666", size = 12),
        axis.title.y = element_text(color = "#666666", size = 12),
        panel.border = element_rect(fill = NA, color = "#666666", size = 0.5, linetype = "solid"),
        panel.grid.major = element_line(size = 0.2, linetype = 'solid', color = "#cccccc"),
        panel.grid.minor = element_line(size = 0.2, linetype = 'solid', color = "#cccccc"))

```

Figure 3 shows historical interest rates (short term, policy rate) in percentages for the US. Monetary policy in the US remained consistent with its objectives and applied tools, for rising consumer prices interest rates stayed high. Data is taken from the [Federal Reserve Economic Data](https://fred.stlouisfed.org/series/IRSTCI01USM156N).

It is clear the interest rate in Turkey has been consistently higher than in the US. The high interest rate environment in Turkey is a result of the recent economic downturns. The US has a more stable economy and can afford to keep interest rates lower.

To see the differences in percentage points clearer, Figure 4 plots the difference between the two interest rates.

### Figure 4:

```{r}
#| echo: false
#| message: false
#| warning: false
# Difference Plot 
# Merge the two data frames by Date
merged_df <- merge(interest_rate_df, interest_rate_df_US, by = "Date", suffixes = c("_TR", "_US"))

merged_df$InterestRate_Diff <- merged_df$InterestRate_TR - merged_df$InterestRate_US

# Plot the difference in interest rates
ggplot(merged_df, aes(x = Date, y = InterestRate_Diff)) +
  geom_line(color = "#1c61b6") +
  ggtitle("Difference in Interest Rates (Turkey - US, Since 2014)") +
  xlab("Date") +
  ylab("Interest Rate Difference (% points)") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, color = "#666666", size = 14, face = "bold"),
        axis.title.x = element_text(color = "#666666", size = 12),
        axis.title.y = element_text(color = "#666666", size = 12),
        panel.border = element_rect(fill = NA, color = "#666666", size = 0.5, linetype = "solid"),
        panel.grid.major = element_line(size = 0.2, linetype = 'solid', color = "#cccccc"),
        panel.grid.minor = element_line(size = 0.2, linetype = 'solid', color = "#cccccc"))
```

Early Period (2014-2019): The difference in interest rates appears relatively stable in the first few years. Turkey's short-term interest rates were consistently higher than the U.S.'s by around 10-12 percentage points, reflecting Turkey's higher inflation and risk profile.

Post-2019 Volatility: From around 2019 onward, the gap becomes more volatile. Large increases and subsequent drops suggest significant changes in Turkish interest rates, which likely coincide with: Central Bank of Turkey interventions:

A response to economic pressures such as inflation and currency depreciation. U.S. Federal Reserve policy shifts: Stable U.S. rates with gradual adjustments, but the volatility here is mostly due to Turkish monetary policy.

Recent Data: The sharp jumps and changes around 2020-2023 suggest a period of significant economic adjustments in Turkey, which may be related to political factors, inflationary pressures, and efforts to stabilize the Turkish lira.

## Inflation in Turkey and in the US

### Figure 5: 

```{r}
#| echo: false
#| message: false
#| warning: false
#| output: false
# Inflation Data TURKEY
symbol_inflation_TR <- "TURCPALTT01CTGYM"

start_date_inflation_TR <- "2014-01-01"  # start of 2014
end_date_inflation_TR <- Sys.Date()  # today

# Get inflation rate data
getSymbols(symbol_inflation_TR, src = "FRED", from = start_date_inflation_TR, to = end_date_inflation_TR)
inflation_rate_TR <- get(symbol_inflation_TR)

# Convert to data frame for ggplot2
inflation_rate_df_TR <- data.frame(Date = index(inflation_rate_TR), InflationRate = as.numeric(inflation_rate_TR))
```

```{r}
#| echo: false
#| message: false
#| warning: false
# Plot inflation rate data
ggplot(inflation_rate_df_TR, aes(x = Date, y = InflationRate)) +
  geom_line(color = "#1c61b6") +
  ggtitle("Turkey Inflation Rate (Since 2014)") +
  xlab("Date") +
  ylab("Inflation Rate (%)") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, color = "#666666", size = 14, face = "bold"),
        axis.title.x = element_text(color = "#666666", size = 12),
        axis.title.y = element_text(color = "#666666", size = 12),
        panel.border = element_rect(fill = NA, color = "#666666", size = 0.5, linetype = "solid"),
        panel.grid.major = element_line(size = 0.2, linetype = 'solid', color = "#cccccc"),
        panel.grid.minor = element_line(size = 0.2, linetype = 'solid', color = "#cccccc"))
```

Figure 4 shows Consumer Price Index change Total for Turkey. In the post-Covid period consumer prices truly skyrocketed, prices increased up to 85% compared to the same month over one year. Data is taken from the [Federal Reserve Economic Data](https://fred.stlouisfed.org/series/TURCPALTT01CTGYM).

### Figure 6: 

```{r}
#| echo: false
#| message: false
#| warning: false
#| output: false
# Inflation Data US
symbol_inflation_US <- "USACPALTT01CTGYM"

start_date_inflation_US <- "2014-01-01"  # start of 2014
end_date_inflation_US <- Sys.Date()  # today

# Get inflation rate data
getSymbols(symbol_inflation_US, src = "FRED", from = start_date_inflation_US, to = end_date_inflation_US)
inflation_rate_US <- get(symbol_inflation_US)

# Convert to data frame for ggplot2
inflation_rate_df_US <- data.frame(Date = index(inflation_rate_US), InflationRate = as.numeric(inflation_rate_US))
```

```{r}
#| echo: false
#| message: false
#| warning: false
# Plot inflation rate data
ggplot(inflation_rate_df_US, aes(x = Date, y = InflationRate)) +
  geom_line(color = "#1c61b6") +
  ggtitle("US Inflation Rate (Since 2014)") +
  xlab("Date") +
  ylab("Inflation Rate (%)") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, color = "#666666", size = 14, face = "bold"),
        axis.title.x = element_text(color = "#666666", size = 12),
        axis.title.y = element_text(color = "#666666", size = 12),
        panel.border = element_rect(fill = NA, color = "#666666", size = 0.5, linetype = "solid"),
        panel.grid.major = element_line(size = 0.2, linetype = 'solid', color = "#cccccc"),
        panel.grid.minor = element_line(size = 0.2, linetype = 'solid', color = "#cccccc"))
```
